Machine Learning Deriven Set of microRNAs as a Novel Biomarker for Myocardial Infarction Diagnosis

### 2023-06-13 Complete-V1/Complete-V1.tex
Revised with Ithenticate and Grammerly

\begin{abstract}
MicroRNAs (miRNAs) play a crucial role in regulating adaptive and
maladaptive responses in cardiovascular diseases, making them attractive
targets for potential therapeutics. However, their potential as novel
biomarkers for diagnosing cardiovascular diseases needs to be evaluated
systematically. In this study, we aimed to identify a key set of miRNA
biomarkers using integrated bioinformatics and machine learning
analysis. We combined and analyzed three gene expression datasets from
the Gene Expression Omnibus (GEO) database, which contained peripheral
blood mononuclear cells samples from individuals with myocardial
infarction (MI), stable coronary artery disease (CAD), and healthy.
Additionally, we selected a set of miRNAs based on their area under the
curve of the receiver operating characteristic (AUC-ROC) for separating
CAD/MI samples. We designed a two-layer architecture for sample
classification, with the first layer isolating healthy samples from
not-healthy ones, and the second layer classifying stable CAD and MI
samples. We trained different machine learning models using both
biomarker sets and evaluated their performance on a test set. We
identified miR-21, miR-186, and miR-32 as the only miRNAs in the
differentially expressed genes, and a set included miR-21, miR-155,
miR-142, miR-197, miR-29A, and miR-320C1 as the optimum set of miRNA
selected by their AUC-ROC. Both biomarker sets were able to distinguish
healthy from not-healthy samples with complete accuracy. The best
performance for classification of CAD and MI was achieved with an SVM
model trained using the biomarker set selected by AUC-ROC, with an AUC
of 0.95 and an accuracy of 0.88. Our study demonstrates that miRNA
signatures derived from peripheral blood could serve as valuable novel
biomarkers for cardiovascular diseases.
\\
\keywords{
        microRNA \and
        Machine Learning \and
        Myocardial Infarction \and
    }

@@

Over the last 50 years, numerous attempts have been collected to use
biomarkers to facilitate diagnosis, assess risk, follow-up therapy, and
determine therapeutic efficacy in CVDs candidates. Based on the released
guidelines, cardiac troponins (cTns) are used as a highly-sensitive and
accurate approach for the detection of myocardial ischemia. Despite the
inherent advantages, the high-rate sensitivity of cTn-based assays has
also led to more false positive results (Thygesen et al. 2018), which do
necessitate the advent and development of new modalities with
pathological values. To improve diagnostic value upon existing CVD
biomarkers, the combination of complementary biological markers, such as
microRNAs (miRNAs) and other genetic factors, is proposed. Previous data
support the notion that miRNAs exhibit the great potential to be used as
alternative biomarkers in CVD detection and follow-up (Schulte et al.
2020). It is suggested that miRNAs possess 18-22 nucleotides and can
play a crucial role in the regulation of gene expression. Evidence point
to the fact that miRNAs are involved in the pathogenesis of cardiac
tissue injury and be as theranostics in terms of CVDs. These elements
can easily circulate in biofluids with both diagnostic as well as
prognostic values (Schulte, Karakas, and Zeller 2017). Several
biological activities such as angiogenesis, cardiomyocyte growth and
contractility, lipid metabolism, plaque formation, and cardiac rhythm
are regulated by miRNAs (Kalayinia et al. 2021). It is postulated that
the function and diagnostic properties of miRNAs are beyond the
myocardium in CVD patients. To be specific, the expression of miRNAs can
vary in different biofluids and cell components such as serum and
peripheral blood mononuclear cells (PBMCs) (Soler-Botija, Gálvez-Montón,
and Bayés-Genís 2019).

PBMCs are a fraction of white blood cells (WBCs), including monocytes,
lymphocytes, macrophages, and other cells belonging to the immune system
(Gao et al. 2020). Emerging data have indicated that PBMCs can be used
as a valid source of biomarkers for monitoring various pathological
conditions. Of note, the alteration of mRNAs and miRNAs under
pathological conditions gives us valuable information about different
kinds of disorders. PBMCs could recapitulate the conditions of the
target tissues, thus, providing a highly sensitive and specific source
of biomarkers (Mosallaei et al. 2022). Commensurate with these
conditions, these cells are repositories of dysregulated genes and
miRNAs expression profiles in CVDs related to control conditions (Gao et
al. 2020; Mosallaei et al. 2022).

@@

Microarray datasets were obtained from the Gene Expression Omnibus (GEO)
database (\url{https://www.ncbi.nlm.nih.gov/geo/}). To obtain sufficient
classification power between MI, healthy and CAD samples, a relatively
large sample size was required. Therefore, GSE59867 for MI and CAD
samples, and GSE56609 and GSE54475 for healthy samples were selected.
All samples were produced using Affymetrix Human Gene 1.0 ST Array
(GPL6244) platform. Only healthy, stable CAD and early-stage MI samples
were selected from these datasets for further analyses. The basic
information for the three datasets evaluated in the current study is
provided in Table @ref(tab:datasets).

@@

For the sake of reaching the highest classification performance using
the set of miRNAs, different models were investigated. To do so, SVM
(with linear, polynomial, and RBF kernels), Logistic Regression (LR),
Random Forests (RF), k-Nearest Neighbor (kNN), Gradient Boosting (GB),
XGBoost (XGB) and Decision Tree (DT) models were trained. All models
were trained with their pre-set parameters with 10-fold
cross-validation. The criteria for choosing the best model were the
highest accuracy and AUC on the test set. The best model was hyper-tuned
with the scikit-opt package (Head et al. 2021) to get the best
predictive performance.

@@

\caption{\label{tab:DEGstab}Total, up-, and down-regulated DEGs and differentially expressed miRNAs.}
\centering
\begin{tabu} to \linewidth {>{\raggedright\arraybackslash}p{3cm}>{\centering}X>{\centering}X>{\centering}X>{\centering\arraybackslash}p{4cm}}
\toprule
  & DEGs & up-regulated DEGs & down-regulated DEGs & miRNAs\\
\midrule
MI vs. Healthy & 860 & 323 & 537 & hsa-miR-186, miR-21, miR-32\\
CAD vs. Healthy & 670 & 262 & 408 & hsa-miR-186, miR-21, miR-32\\
MI vs. CAD & 260 & 144 & 116 & hsa-miR-186\\
\bottomrule
\end{tabu}
\end{table}

@@

{\centering \includegraphics[width=0.6\linewidth]{../../DEG and DEM/Pics/venny}

@@

\begin{figure}
\includegraphics[width=0.95\linewidth]{../../DEG and DEM/Pics/enrich MI-Healthy} \caption{Gene Ontology (GO) and Kyoto Encyclopedia of Genes and Genomes (KEGG) pathways enriched with the MI and healthy DEGs. (A) Biological process terms. (B) Cellular component terms. (C) Molecular function terms. (D) KEGG analysis.}\label{fig:MIHEnrich}
\end{figure}

@@

\begin{figure}
\includegraphics[width=0.95\linewidth]{../../DEG and DEM/Pics/enrich CAD-Healthy} \caption{Gene Ontology (GO) and Kyoto Encyclopedia of Genes and Genomes (KEGG) pathways enriched with the CAD and healthy DEGs. (A) Biological process terms. (B) Cellular component terms. (C) Molecular function terms. (D) KEGG analysis.}\label{fig:CADHEnrich}
\end{figure}

@@

Among all DEGs, just miR-186, miR-32, and miR-21 were detected as
differentially expressed miRNAs. The expression profile of these three
miRNAs is presented in Figure @ref(fig:DEMexp). Additionally, The ROC
curves of each miRNA for each layer are presented in Figure
@ref(fig:miRROC). Using the logistic regression model the AUC for
miR-21, miR-32, and miR-186 was 0.99, 1, and 0.91 respectively (Figure
@ref(fig:miRROC)A). Besides, the accuracy of each miRNA for classifying
the samples into healthy and not-healthy groups on the test set was
0.92, 0.98, and 0.89 for miR-21, miR-32, and miR-186, respectively.
Moreover, the ROC curve of each miRNA for classifying MI and CAD samples
was presented in Figure @ref(fig:miRROC)B. For miR-21, miR-32, and
miR-186, the AUC and accuracy on the test set were 0.85; 0.7; and 0.82,
and 0.78; 0.67; and 0.74, respectively.

@@

\caption{\label{tab:mirExptable}Target miRNAs log fold-change and adjusted p-values for CAD samples relative to healthy, MI samples relative to healthy, and MI samples relative to CAD.}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{CAD/Healthy} & \multicolumn{2}{c}{MI/Healthy} & \multicolumn{2}{c}{MI/CAD} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
  & logFC & adj. p-value & logFC & adj. p-value & logFC & adj. p-value\\
\midrule
miR-186 & 1.4 & 3.60e-25 & 0.9 & 6.76e-20 & -0.5 & 1.05e-09\\
miR-21 & 1.4 & 1.31e-17 & 2.3 & 2.07e-47 & 0.8 & 2.96e-11\\
miR-32 & 2.5 & 8.39e-43 & 2.2 & 3.10e-59 & -0.3 & 7.60e-04\\
miR-155 & -0.6 & 2.49e-12 & -0.9 & 7.59e-33 & -0.2 & 2.68e-07\\
miR-142 & 0.2 & 1.90e-01 & -0.1 & 1.70e-01 & -0.3 & 2.90e-04\\
miR-197 & 0.5 & 2.95e-20 & 0.7 & 1.59e-47 & 0.2 & 8.58e-09\\
miR-29A & 0.7 & 7.76e-29 & 0.1 & 1.70e-01 & -0.5 & 2.14e-10\\
miR-320C1 & 0.8 & 4.72e-22 & 0.5 & 1.89e-30 & -0.2 & 1.84e-05\\
\bottomrule
\end{tabu}
\end{table}

@@

{\centering \includegraphics[width=0.5\linewidth]{../../ML/Pics/DEGs healthy not-healthy/DEMs Expression}

@@

\caption{Expression profile of miR-186, miR-21, and miR32 in Healthy, CAD, and MI samples.}\label{fig:DEMexp}
\end{figure}

@@

Although single miRNAs had acceptable performance, their predictive
value could be improved even further by using them as a set. The ROC
curve for the SVM model with an RBF kernel trained with all three miRNAs
is presented in Figure @ref(fig:DEMROC)A. The model had a better
performance in classification than single miRNAs.The AUC for the model
is 1, and its accuracy on the test set was also 1. In Figure
@ref(fig:DEMCF)A, the confusion matrix for the model is presented.

@@

Different models were trained using expression values for three
differentially miRNAs. The models' 10-fold cross-validated AUC and
accuracy on the test set are reported in Figure @ref(fig:DEMmodels). The
best model from both AUC and accuracy point-of-view was the SVM model
with linear kernel. The AUC and accuracy for this model with its pre-set
values were 0.93 and 0.82 respectively. The model was hyper-tuned for C
and gamma hyper-parameters, and therefore the model showed better
performance. The ROC curve of the hyper-tuned model is presented in
Figure @ref(fig:DEMROC)B. For this model the AUC reached 0.95 and the
accuracy improved to 0.85 (Table @ref(tab:DEGsML)). Moreover, the
sensitivity and specificity for the model on the test set were 0.91 and
0.71 respectively. The confusion matrix for the hyper-tuned model is
illustrated in Figure @ref(fig:DEMCF)B.

@@

\caption{Area under the curve (AUC) and accuracy of different models trained with three miRNAs in DEGs on the test set.}\label{fig:DEMmodels}
\end{figure}

@@

\caption{\label{tab:DEGsML}AUC-ROC and accuracy for SVM with the linear kernel as the best model trained with differentially expressed miRNAs on the train and test set before and after hyper-tuning}
\centering
\begin{tabu} to \linewidth {>{\centering}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{2}{c}{Pre-set parameters} & \multicolumn{2}{c}{Hyper-tuned} \\
\cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6}
Model & Metrics & train & test & train & test\\
\midrule
SVM & AUC-ROC & 0.91 & 0.93 & 0.92 & 0.95\\
(Linear Kernel) & Accuracy & 0.83 & 0.82 & 0.84 & 0.85\\
\bottomrule
\end{tabu}
\end{table}

\hypertarget{auc-approach}{%
\subsubsection{AUC approach}\label{auc-approach}}

After calculating the AUC for each miRNA for the classification of MI
and CAD samples, miRNAs were sorted, and their performance was
investigated as a set. The metric of choice for selecting the best set
was AUC. As shown in Figure @ref(fig:AUCset), the AUC increased until
the number of miRNAs in the set reached six, and after that, it dropped.
The AUC for separating MI samples from CAD using these miRNAs was 0.93.
The miRNAs in the set were miR-29a, miR-197, miR-142, miR-21, miR-155,
and miR-320C1. The expression values of these miRNAs in healthy, CAD,
and MI samples are presented in Figure @ref(fig:AUCexp). The ROC curve
of the selected miRNAs for MI and CAD sample classification is
illustrated in Figure @ref(fig:miRROC)B.

{\centering \includegraphics[width=0.65\linewidth]{../../ML/Pics/AUC for cad mi mirs}

\caption{Area under the curve (AUC) for sets containing an increasing number of miRNAs with the highest individual AUC in MI/CAD separation.}\label{fig:AUCset}
\end{figure}

{\centering \includegraphics[width=0.9\linewidth]{../../ML/Pics/ExpressionforAUC}

\caption{Expression profile of has-miR-29A, has-miR-197, has-miR-142, has-miR-21, has-miR-155, and has-miR-320C1 in Healthy, CAD, and MI samples.}\label{fig:AUCexp}
\end{figure}

@@

Using the selected set, an SVM model with an RBF kernel was trained to
separate healthy from not-healthy samples. The ROC curve for the model
is presented in Figure @ref(fig:AUCROC)A and the confusion matrix is
illustrated in Figure @ref(fig:AUCCF)A. Both AUC and accuracy for the
model on the test set were equal to 1.

@@

\caption{ROC curve for the set of miRNAs selected by AUC on test set classification. (A) SVM with RBF kernel for healthy and not-healthy samples classification. (B) Logistic regression model for CAD and MI samples classification. (C) SVM with polynomial kernel for CAD and MI samples classification. }\label{fig:AUCROC}
\end{figure}

@@

To find the best model for training the best set, different models were
trained using their pre-set values. Their AUC and accuracy results on
the test set are presented in Figure @ref(fig:AUCmodels). The best model
from the AUC point-of-view was the LR and from the accuracy
point-of-view, it was the SVM model with a polynomial kernel. For the LR
model the AUC and accuracy were 0.92 and 0.81, respectively; and for the
SVM model with a polynomial kernel, the values were 0.91 and 0.84,
respectively. Both models were hyper-tuned and the ROC curve for their
best performance is presented in Figure @ref(fig:AUCROC)B and C. The AUC
and accuracy for the LR model increased to 0.94 and 0.88, respectively.
For the SVM model with a polynomial kernel, these values increased to
0.95 and 0.88, respectively (Table @ref(tab:AUCML)). The sensitivity for
the LR and SVM models were 1 and 0.97, respectively; and the specificity
for them was 0.57 and 0.64, respectively. The confusion matrix for both
models is illustrated in Figure @ref(fig:AUCCF)B and C.

@@

{\centering \includegraphics[width=0.65\linewidth]{../../ML/Pics/Models AUC mirs}

@@

\caption{Area under the curve (AUC) and accuracy of different models trained with three miRNAs in DEGs on the test set.}\label{fig:AUCmodels}
\end{figure}

@@

\caption{\label{tab:AUCML}AUC-ROC and accuracy for SVM with the linear kernel as the best model trained with miRNAs selected based on their individual AUC-ROC on the train and test set before and after hyper-tuning}
\centering
\begin{tabu} to \linewidth {>{\centering}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{2}{c}{Pre-set parameters} & \multicolumn{2}{c}{Hyper-tuned} \\
\cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6}
Model & Metrics & Train & Test & Train & Test\\
\midrule
LR & AUC-ROC & 0.90 & 0.92 & 0.91 & 0.94\\
 & Accuracy & 0.84 & 0.81 & 0.84 & 0.88\\
SVM & AUC-ROC & 0.90 & 0.91 & 0.91 & 0.95\\
(Polynomial Kernel) & Accuracy & 0.86 & 0.84 & 0.86 & 0.88\\
\bottomrule
\end{tabu}
\end{table}

@@

The prevalence of acute MI (AMI) can lead to high-rate mortality in the
clinical setting. However, early diagnosis and application of suitable
treatment protocols can reduce mortality and improve AMI prognosis
({``Cardiovascular Diseases ({CVDs})''} n.d.; Thygesen et al. 2018; Tsao
et al. 2022). Studies have suggested that changes in miRNA expression
may play a significant role in the progression of MI and the subsequent
remodeling (Laggerbauer and Engelhardt 2022). It is believed that the
expression of miRNAs is altered during the various biological processes
correlated with MI within the myocardium or other related tissues (Khan,
Gupta, and Mahapatra 2022). Although several research has been
concentrated on examining free circulating miRNAs in the serum samples
for the detection of cardiac tissue injuries (Kaur et al. 2020), more
information is needed to fully comprehend the miRNAs found in different
blood sub-components like plasma, platelets, and PBMCs. Based on
previous data, PBMCs are critically involved in plaque destabilization
and rupture as well as early inflammatory responses during MI (Mosallaei
et al. 2022; Hapke et al. 2022). Moreover, PBMCs have specific miRNA
profile that is altered under certain pathological conditions which are
great candidates as disease biomarkers (Mosallaei et al. 2022).

@@

In summary, we derived a set of miRNA biomarkers by comparing MI samples
to both healthy and CAD samples. We found that the SVM model performed
best in both the first layer, which separated healthy and not-healthy
samples, and the second layer, which classified MI/CAD samples. The set
of miRNAs selected based on their AUC values had slightly better
performance in the second layer. Overall, our two-layer structure
achieved a weighted accuracy of 0.91. This demonstrates the potential
for combining bioinformatics and machine learning techniques to identify
novel biomarkers and gain a deeper understanding of myocardial
infarction.

>>>>

\begin{abstract}
MicroRNAs (miRNAs) play a crucial role in regulating adaptive and
maladaptive responses in cardiovascular diseases, making them attractive
targets for potential therapeutics. However, their potential as novel
biomarkers for diagnosing cardiovascular diseases requires systematic
evaluation. In this study, we aimed to identify a key set of miRNA
biomarkers using integrated bioinformatics and machine learning
analysis. We combined and analyzed three gene expression datasets from
the Gene Expression Omnibus (GEO) database, which contains peripheral
blood mononuclear cell samples from individuals with myocardial
infarction (MI), stable coronary artery disease (CAD), and healthy
individuals. Additionally, we selected a set of miRNAs based on their
area under the curve of the receiver operating characteristic curve
(AUC-ROC) to separate the CAD/MI samples. We designed a two-layer
architecture for sample classification, with the first layer isolating
healthy samples from not-healthy ones, and the second layer classifying
stable CAD and MI samples. We trained different machine learning models
using both biomarker sets and evaluated their performance on a test set.
We identified miR-21, miR-186, and miR-32 as the only miRNAs among the
differentially expressed genes, and a set including miR-186, miR-21,
miR-197, miR-29A, and miR-296 as the optimum set of miRNAs selected by
their AUC-ROC. Both biomarker sets could distinguish healthy from
not-healthy samples with complete accuracy. The best performance for the
classification of CAD and MI was achieved with an SVM model trained
using the biomarker set selected by AUC-ROC, with an AUC-ROC of 0.95,
and an accuracy of 0.88. Our study demonstrated that miRNA signatures
derived from peripheral blood could serve as valuable novel biomarkers
for cardiovascular diseases.
\\
\keywords{
        microRNA \and
        Machine Learning \and
        Myocardial Infarction \and
    }

@@

Over the last 50 years, numerous attempts have been collected to use
biomarkers to facilitate diagnosis, assess risk, follow-up therapy, and
determine therapeutic efficacy in CVDs candidates. Based on the released
guidelines, cardiac troponins (cTns) are used as a highly-sensitive and
accurate approach for the detection of MI. Despite the inherent
advantages, the high-rate sensitivity of cTn-based assays has also led
to more false positive results (Thygesen et al. 2018), which do
necessitate the advent and development of new modalities with
pathological values. To improve diagnostic value upon existing MI
biomarkers, the combination of complementary biological markers, such as
microRNAs (miRNAs) and other genetic factors, is proposed. Previous
researches support the notion that miRNAs exhibit the great potential to
be used as alternative biomarkers in CVDs detection and follow-up
(Schulte et al. 2020). It is suggested that miRNAs possess 18-22
nucleotides and can play a crucial role in the regulation of gene
expression. Evidence point to the fact that miRNAs are involved in the
pathogenesis of cardiac tissue injury and could be considered as
theranostics targets in terms of CVDs. These elements can easily
circulate in biofluids with both diagnostic as well as prognostic values
(Schulte, Karakas, and Zeller 2017). Several biological activities such
as angiogenesis, cardiomyocyte growth and contractility, lipid
metabolism, plaque formation, and cardiac rhythm are regulated by miRNAs
(Kalayinia et al. 2021). It is postulated that the function and
diagnostic properties of miRNAs are beyond the myocardium in CVD
patients. To be specific, the expression of miRNAs can vary in different
biofluids and cell components such as serum and peripheral blood
mononuclear cells (PBMCs) (Soler-Botija, Gálvez-Montón, and Bayés-Genís
2019).

PBMCs are a fraction of white blood cells (WBCs), including monocytes,
lymphocytes, macrophages, and other cells belonging to the immune system
(Gao et al. 2020). Emerging data have indicated that PBMCs can be used
as a valid source of biomarkers for monitoring various pathological
conditions. Of note, the alteration of mRNAs and miRNAs under
pathological conditions gives us valuable information about different
kinds of disorders. PBMCs could recapitulate the conditions of the
target tissues, thus, providing a highly sensitive and specific source
of biomarkers (Mosallaei et al. 2022). Commensurate with these
conditions, these cells are repositories of dysregulated genes and
miRNAs expression profiles in CVDs (Gao et al. 2020; Mosallaei et al.
2022).

@@

Microarray datasets were obtained from the Gene Expression Omnibus (GEO)
database (\url{https://www.ncbi.nlm.nih.gov/geo/}). To obtain sufficient
classification power between MI, healthy and stable coronary artery
disease (CAD) samples, a relatively large sample size was required.
Therefore, GSE59867 for MI and CAD samples, and GSE56609 and GSE54475
for healthy samples were selected. All samples were produced using
Affymetrix Human Gene 1.0 ST Array (GPL6244) platform. Only healthy,
stable CAD and early-stage MI samples were selected from these datasets
for further analyses. The basic information for the three datasets
evaluated in the current study is provided in Table @ref(tab:datasets).

@@

For the sake of reaching the highest classification performance using
the set of miRNAs, different models were investigated. To do so, SVM
(with linear, polynomial, and RBF kernels), Logistic Regression (LR),
Random Forests (RF), k-Nearest Neighbor (kNN), Gradient Boosting (GB),
XGBoost (XGB) and Decision Tree (DT) models were trained. All models
were trained with their pre-set parameters with 10-fold
cross-validation. The criteria for choosing the best model were the
highest accuracy and AUC-ROC on the test set. The best model was
hyper-tuned with the scikit-opt package (Head et al. 2021) to get the
best predictive performance.

@@

\caption{\label{tab:DEGstab}Total, up-, and down-regulated DEGs and differentially expressed miRNAs.}
\centering
\begin{tabu} to \linewidth {>{\raggedright\arraybackslash}p{3cm}>{\centering}X>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{5cm}}
\toprule
  & Total DEGs & Up-regulated DEGs & Down-regulated DEGs & miRNAs\\
\midrule
MI vs. Healthy & 860 & 323 & 537 & hsa-miR-186, miR-21, miR-32\\
CAD vs. Healthy & 670 & 262 & 408 & hsa-miR-186, miR-21, miR-32\\
MI vs. CAD & 260 & 144 & 116 & hsa-miR-186\\
\bottomrule
\end{tabu}
\end{table}

@@

{\centering \includegraphics[width=0.45\linewidth]{../../DEG and DEM/Pics/venny}

@@

\begin{figure}
\includegraphics[width=0.95\linewidth]{../../DEG and DEM/Pics/Enrichment MI-Healthy} \caption{Gene Ontology (GO) and Kyoto Encyclopedia of Genes and Genomes (KEGG) pathways enriched with the MI and healthy DEGs. (A) Biological process terms. (B) Cellular component terms. (C) Molecular function terms. (D) KEGG analysis.}\label{fig:MIHEnrich}
\end{figure}

@@

\begin{figure}
\includegraphics[width=0.95\linewidth]{../../DEG and DEM/Pics/Enrichment CAD-Healthy} \caption{Gene Ontology (GO) and Kyoto Encyclopedia of Genes and Genomes (KEGG) pathways enriched with the CAD and healthy DEGs. (A) Biological process terms. (B) Cellular component terms. (C) Molecular function terms. (D) KEGG analysis.}\label{fig:CADHEnrich}
\end{figure}

@@

Among all DEGs, just miR-186, miR-32, and miR-21 were detected as
differentially expressed miRNAs. The expression profile of these three
miRNAs is presented in Figure @ref(fig:Expall). Additionally, The ROC
curves of each miRNA for each layer are presented in Figure
@ref(fig:miRROC). Using the logistic regression model the AUC-ROC of
miR-21, miR-32, and miR-186 for separating healthy and not-healthy
samples was 0.98, 0.99, and 0.90 respectively (Figure
@ref(fig:miRROC)A). Besides, the accuracy of each miRNA for classifying
the samples into healthy and not-healthy groups on the test set was
0.92, 0.98, and 0.89 for miR-21, miR-32, and miR-186, respectively.
Moreover, the ROC curve of each miRNA for classifying MI and CAD samples
was presented in Figure @ref(fig:miRROC)B. For miR-21, miR-32, and
miR-186, the AUC-ROC and accuracy on the test set were 0.85; 0.70; and
0.86, and 0.78; 0.67; and 0.74, respectively.

@@

\caption{\label{tab:mirExptable}Target miRNAs log fold-change and adjusted p-values for CAD samples relative to healthy, MI samples relative to healthy, and MI samples relative to CAD.}
\centering
\begin{tabu} to \linewidth {>{\raggedright}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{CAD/Healthy} & \multicolumn{2}{c}{MI/Healthy} & \multicolumn{2}{c}{MI/CAD} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
  & logFC & adj. p-value & logFC & adj. p-value & logFC & adj. p-value\\
\midrule
miR-186 & 1.4 & 3.60e-25 & 0.9 & 6.76e-20 & -0.5 & 1.05e-09\\
miR-21 & 1.4 & 1.31e-17 & 2.3 & 2.07e-47 & 0.8 & 2.96e-11\\
miR-32 & 2.5 & 8.39e-43 & 2.2 & 3.10e-59 & -0.3 & 7.60e-04\\
miR-197 & 0.5 & 2.95e-20 & 0.7 & 1.59e-47 & 0.2 & 8.58e-09\\
miR-29A & 0.7 & 7.76e-29 & 0.1 & 1.70e-01 & -0.5 & 2.14e-10\\
miR-296 & -0.1 & 5.00e-02 & 0.1 & 2.00e-02 & 0.2 & 6.15e-06\\
\bottomrule
\end{tabu}
\end{table}

@@

{\centering \includegraphics[width=0.9\linewidth]{../../ML/Pics/Expressionforall}

@@

\caption{Expression profile of all miRNAs in two approach in different sample classes.}\label{fig:Expall}
\end{figure}

@@

Although single miRNAs had acceptable performance, their predictive
value could be improved even further by using them as a set. The ROC
curve for the SVM model with an RBF kernel trained with all three miRNAs
is presented in Figure @ref(fig:DEMROC)A. The model had a better
performance in classification than single miRNAs.The AUC-ROC for the
model is 1, and its accuracy on the test set was also 1. In Figure
@ref(fig:DEMCF)A, the confusion matrix for the model is presented.

@@

Different models were trained using expression values for three
differentially miRNAs. The models' 10-fold cross-validated AUC-ROC and
accuracy on the test set are reported in Figure @ref(fig:DEMmodels). The
best model from both AUC-ROC and accuracy point-of-view was the SVM
model with linear kernel. The AUC-ROC and accuracy for this model with
its pre-set values were 0.93 and 0.82 respectively. The model was
hyper-tuned for C and gamma hyper-parameters, and therefore the model
showed better performance. The ROC curve of the hyper-tuned model is
presented in Figure @ref(fig:DEMROC)B. For this model the AUC-ROC
reached 0.95 and the accuracy improved to 0.85 (Table @ref(tab:DEGsML)).
Moreover, the sensitivity and specificity for the model on the test set
were 0.91 and 0.71 respectively. The confusion matrix for the
hyper-tuned model is illustrated in Figure @ref(fig:DEMCF)B.

@@

\caption{Area under the receiver operating characteristic curve and accuracy of different models trained with three miRNAs in DEGs on the test set.}\label{fig:DEMmodels}
\end{figure}

@@

\caption{\label{tab:DEGsML}AUC-ROC and accuracy for SVM with the linear kernel as the best model trained with differentially expressed miRNAs on the train and test set before and after hyper-tuning}
\centering
\begin{tabu} to \linewidth {>{\centering}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{2}{c}{Pre-set parameters} & \multicolumn{2}{c}{Hyper-tuned} \\
\cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6}
Model & Metrics & train & test & train & test\\
\midrule
SVM-linear & AUC-ROC & 0.91 & 0.93 & 0.92 & 0.95\\
 & Accuracy & 0.83 & 0.82 & 0.84 & 0.85\\
\bottomrule
\end{tabu}
\end{table}

\hypertarget{auc-roc-approach}{%
\subsubsection{AUC-ROC approach}\label{auc-roc-approach}}

After calculating the AUC-ROC for each miRNA for the classification of
MI and CAD samples, the miRNAs with AUC-ROC over 0.8 were selected. The
selected miRNAs were miR-29a, miR-197, miR-186, miR-21, and miR-296. The
expression values of these miRNAs in healthy, CAD, and MI samples are
presented in Figure @ref(fig:Expall). The ROC curve of the selected
miRNAs for both layers are illustrated in Figure @ref(fig:miRROC).

@@

Using the selected set, an SVM model with an RBF kernel was trained to
separate healthy from not-healthy samples. The ROC curve for the model
is presented in Figure @ref(fig:AUCROC)A and the confusion matrix is
illustrated in Figure @ref(fig:AUCCF)A. Both AUC-ROCand accuracy for the
model on the test set were equal to 1.

@@

\caption{ROC curve for the set of miRNAs selected by AUC-ROC on test set classification. (A) SVM with RBF kernel for healthy and not-healthy samples classification. (B) Logistic regression model for CAD and MI samples classification. (C) SVM with polynomial kernel for CAD and MI samples classification. }\label{fig:AUCROC}
\end{figure}

@@

To find the best model for training the best set, different models were
trained using their pre-set values. Their AUC-ROC and accuracy results
on the test set are presented in Figure @ref(fig:AUCmodels). The best
model from the AUC-ROC point-of-view was the SVM with linear kernel and
from the accuracy point-of-view, it was the SVM model with an RBF
kernel. For the SVM-linear model the AUC-ROC and accuracy were 0.93 and
0.82, respectively; and for the SVM-RBF, the values were 0.92 and 0.84,
respectively. Both models were hyper-tuned and the ROC curve for their
best performance is presented in Figure @ref(fig:AUCROC)B and C. The
AUC-ROC and accuracy for the SVM-linear model modified to 0.92 and 0.88,
respectively. For the SVM-RBF, these values increased to 0.96 and 0.94,
respectively (Table @ref(tab:AUCML)). The sensitivity for the SVM-linear
and SVM-RBF models were 0.91 and 0.97, respectively; and the specificity
for them was 0.79 and 0.86, respectively. The confusion matrix for both
models is illustrated in Figure @ref(fig:AUCCF)B and C.

@@

{\centering \includegraphics[width=0.65\linewidth]{../../ML/Pics/Models AUC 5 mirs CV}

@@

\caption{Area under the receiver operating characteristic curve and accuracy of different models trained with AUC-selected miRNAs.}\label{fig:AUCmodels}
\end{figure}

@@

\caption{\label{tab:AUCML}AUC-ROC and accuracy for SVM with the linear kernel as the best model trained with miRNAs selected based on their individual AUC-ROC on the train and test set before and after hyper-tuning}
\centering
\begin{tabu} to \linewidth {>{\centering}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X>{\centering}X}
\toprule
\multicolumn{2}{c}{ } & \multicolumn{2}{c}{Pre-set parameters} & \multicolumn{2}{c}{Hyper-tuned} \\
\cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6}
Model & Metrics & Train & Test & Train & Test\\
\midrule
SVM-linear & AUC-ROC & 0.91 & 0.93 & 0.93 & 0.92\\
 & Accuracy & 0.85 & 0.82 & 0.90 & 0.88\\
SVM-RBF & AUC-ROC & 0.90 & 0.92 & 0.96 & 0.96\\
 & Accuracy & 0.86 & 0.84 & 0.96 & 0.94\\
\bottomrule
\end{tabu}
\end{table}

@@

The prevalence of MI can lead to high-rate mortality in the clinical
setting. However, early diagnosis and application of suitable treatment
protocols can reduce mortality and improve AMI prognosis
({``Cardiovascular Diseases ({CVDs})''} n.d.; Thygesen et al. 2018; Tsao
et al. 2022). Studies have suggested that changes in miRNA expression
may play a significant role in the progression of MI and the subsequent
remodeling (Laggerbauer and Engelhardt 2022). It is believed that the
expression of miRNAs is altered during the various biological processes
correlated with MI within the myocardium or other related tissues (Khan,
Gupta, and Mahapatra 2022). Although several research has been
concentrated on examining free circulating miRNAs in the serum samples
for the detection of cardiac tissue injuries (Kaur et al. 2020), more
information is needed to fully comprehend the miRNAs found in different
blood sub-components like plasma, platelets, and PBMCs. Based on
previous data, PBMCs are critically involved in plaque destabilization
and rupture as well as early inflammatory responses during MI (Mosallaei
et al. 2022; Hapke et al. 2022). Moreover, PBMCs have specific miRNA
profile that is altered under certain pathological conditions which are
great candidates as disease biomarkers (Mosallaei et al. 2022).

@@

In summary, we derived a set of miRNA biomarkers by comparing MI samples
to both healthy and CAD samples. We found that the SVM model performed
best in both the first layer, which separated healthy and not-healthy
samples, and the second layer, which classified MI/CAD samples. The set
of miRNAs selected based on their AUC-ROC values had better performance
in the second layer. Overall, our two-layer structure achieved a
accuracy of 0.96. This demonstrates the potential for combining
bioinformatics and machine learning techniques to identify novel
biomarkers and gain a deeper understanding of myocardial infarction.
